---
title: "Customizing Charts Using ggplot2 extensions"
author: "Cyrus Wakibia"
date: "2023-09-24"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---


*This document was last updated on `r format(Sys.Date(), "%Y-%m-%d")`.*


```{r load packages, include=FALSE}
library(tidyverse)
library(ggtext)
library(glue)
library(ggforce)
library(ggrepel)
library(geomtextpath)

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```

### Some important themes

- `theme_minimal()`
- `ggthemes`
- `bbplots`
- `ggthemr`
- `The Economist`

## ggthemes: Improve your themes with this

# gghighlight: Highlight specific parts of the chart

# ggtext

Provides simple Markdown and HTML rendering for ggplot2

#### [reference](https://github.com/wilkelab/ggtext/)

### Key Functions

```         
- element_markdown()
- element_textbox()
```

**`element_markdown()`**

-   Both behave similarly to element_text() but render the provided text as markdown/html. `element_markdown()` is meant as a direct replacement for *`element_text()`*, and it renders text without word wrapping.
-   To start a new line, use the **`<br>` tag or add two spaces** before the end of a line.

**`element_textbox()`**

-   `element_textbox()` offers support for rendering larger amounts of text that require word wrapping.
-   Unlike `element_markdown()`, it cannot be used for axis tick labels, and it cannot draw text at arbitrary angles, only at fixed orientations corresponding to 0, 90, 180, and 270 degrees.
-   In practice, you will usually want to use *`element_textbox_simple()`* instead of *`element_textbox()`*, as it sets useful defaults for many parameters not usually defined in ggplot2 themes.

#### element_markdown()

```{r}

data("diamonds")
color = c("#009E73", "#D55E00", "#0072B2", "#000000", "#F11F00")


diamonds |> 
  group_by(cut) |> 
  summarise(average_price = mean(price)) |> 
  mutate(
         cut = glue("<b style='color:{color}'>{cut}</b>"),
         cut = fct_reorder(cut, average_price)
         ) |> 
   ggplot(aes(average_price, cut, fill = color)) + 
  geom_col(alpha = 0.5) + 
  scale_fill_identity() +
  
  ## Add the markdown styling
  labs(caption = "Data Source: **ggplot2**<br>",
       title = "**Diamond's *Average Price* by *CUT* type**") +
  theme(
    axis.text.y = element_markdown(),
    plot.caption = element_markdown(lineheight = 1.2),
    plot.title = element_markdown()
  )


```

#### element_markdown()

```{r}
ggplot(mtcars, aes(disp, mpg)) + 
  geom_point() +
  labs(
    title = "<b>Fuel economy vs. engine displacement</b><br>
    <span style = 'font-size:10pt'>There is a *negative*
    linear **relationship** between <span style = 'color:red;'>mpg and discplacement</span>.</span>",
    x = "displacement (in<sup>3</sup>)",
    y = "Miles per gallon (mpg)<br><span style = 'font-size:8pt'>A measure of
    the car's fuel efficiency.</span>"
  ) +
  theme(
    plot.title.position = "plot",
    plot.title = element_textbox_simple(
      size = 13,
      lineheight = 1,
      padding = margin(5.5, 5.5, 5.5, 5.5),
      margin = margin(0, 0, 5.5, 0),
      fill = "cornsilk"
    ),
    axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      hjust = 0,
      orientation = "left-rotated",
      minwidth = unit(1, "in"),
      maxwidth = unit(2, "in"),
      padding = margin(4, 4, 2, 4),
      margin = margin(0, 0, 2, 0),
      fill = "lightsteelblue1"
    )
  )
```

# ggrepel

Repel overlapping text labels away from each other.

It provides the following geoms: - geom_text_repel(): adds text directly to the plot - geom_label_repel(): draws a rectangle underneath the text, making it easier to read

```         
**We can also disable clipping to allow the labels to go beyond the edges of the panel.**
```

**geom_text_repel**

*Text labels repel away from each other, away from data points, and away from edges of the plotting area (panel).*

```{r}
ggplot(mtcars, aes(wt, mpg, label = rownames(mtcars))) +
  geom_text_repel() +
  geom_point(color = 'red') +
  theme_classic(base_size = 16)
```

```{r}
set.seed(42)

p <- ggplot(mtcars, aes(y = wt, x = 1, label = rownames(mtcars))) +
  geom_point(color = "red") +
  ylim(1, 5.5) +
  theme(
    axis.line.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.title.x = element_blank()
  )

p1 <- p +
  xlim(1, 1.375) +
  geom_text_repel(
    force        = 0.5,
    nudge_x      = 0.15,
    direction    = "y",
    hjust        = 0,
    segment.size = 0.2
  ) +
  ggtitle("hjust = 0")

p2 <- p + 
  xlim(1, 1.375) +
  geom_text_repel(
    force        = 0.5,
    nudge_x      = 0.2,
    direction    = "y",
    hjust        = 0.5,
    segment.size = 0.2
  ) +
  ggtitle("hjust = 0.5 (default)")

p3 <- p +
  xlim(0.25, 1) +
  scale_y_continuous(position = "right") +
  geom_text_repel(
    force        = 0.5,
    nudge_x      = -0.25,
    direction    = "y",
    hjust        = 1,
    segment.size = 0.2
  ) +
  ggtitle("hjust = 1")

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

**geom_label_repel**

```{r}
p <- ggplot(mtcars,
  aes(wt, mpg, label = rownames(mtcars), colour = factor(cyl))) +
  geom_point()
# Labels with background
p + geom_label_repel()
```

# ggforce

-   The `ggforce` package is a ggplot2 extension that adds many exploratory data analysis features

    hull plots for visualizing clusters or groups within our data

-   The Hull Plot is a visualization that produces a shaded areas around clusters (groups) within our data. It gets the name because of the Convex Hull shape. It's a great way to show customer segments, group membership, and clusters on a Scatter Plot.

```{r}
library(tidyquant)

mpg %>%
  mutate(engine_size = str_c("Cylinder: ", cyl)) %>%
  ggplot(aes(displ, hwy)) +
  geom_mark_ellipse(
        aes(fill = engine_size, label = engine_size),
        label.buffer = unit(4, 'mm'),
        label.fill = 'tomato',
        show.legend = FALSE)+
  geom_point()+
  geom_smooth(se = FALSE, span = 1.0, method = 'loess') +
  expand_limits(y = 50) +
  theme_tq() +
  scale_fill_tq() +
  labs(
      title = "Fuel Economy (MPG) Trends by Engine Size and Displacement",
      subtitle = "Hull plot to indicate clusters / group assignment",
      y = "Highway Fuel Economy (MPG)",
      x = "Engine Displacement Volume (Liters)",
      fill = "",
      caption = "Engine size has a negative relationship to fuel economy."
  )





```


### ggplot2 recap

- `geom_path()` connects the observations in the order in which they appear in the data. 
- `geom_line()` connects them in order of the variable on the x axis. 
- `geom_step()` creates a stairstep plot, highlighting exactly when changes occur. The group aesthetic determines which cases are connected together.

#### `geom_line`

```{r}
recent <- economics[economics$date > as.Date("2013-01-01"), ]
ggplot(recent, aes(date, unemploy)) + geom_line()

```

#### `geom_step`

```{r}

ggplot(recent, aes(date, unemploy)) + geom_step()

```

#### `geom_path`

```{r}
m <- ggplot(economics, aes(unemploy/pop, psavert))
m + geom_path()
# m + geom_path(aes(colour = as.numeric(date)))

```


# geomtextpath

[reference](https://allancameron.github.io/geomtextpath/)


- The goal of the `geomtextpath` package is to directly label line-based plots with text that is able to follow a curved path
-   Using `geomtextpath`, our text can follow any path, and will remain correctly spaced and angled, even if you change the size and aspect ratio of your plotting device.

-   The core functions in this package(`geomtextpath`), `geom_textpath` and `geom_labelpath`, work like any other geom in ggplot2.

-   They take their co-ordinates, their y co-ordinates and their text label from an aesthetic mapping.


    - The geoms for path, segment, *line, curve, density*, smooth, contour, sf; have two corresponding alternatiaves. e.g., textline and labelline. This follows for each of them
    

```{r}
# Set a consistent theme for the plots here
theme_set(theme_minimal() + 
          theme(axis.line = element_line(size = 0.25, colour = "gray75")))

t <- seq(5, -1, length.out = 1000) * pi

spiral <- data.frame(x    = sin(t) * 1:1000, 
                     y    = cos(t) * 1:1000,
                     text = paste("Like a circle in a spiral,",
                                  "like a wheel within a wheel,",
                                  "never ending or beginning,",
                                  "on an ever spinning reel")
                     )

ggplot(spiral, aes(x, y, label = text)) +
  geom_textpath(linewidth = 7, vjust = 2, text_only = TRUE) +
  coord_equal(xlim = c(-1500, 1500), ylim = c(-1500, 1500))
```

Add text in a textbox

```{r}

set.seed(5)

df <- data.frame(x = spline(1:5, runif(5), xout = seq(1, 5, 1/100))$y,
                 y = spline(1:5, runif(5), xout = seq(1, 5, 1/100))$y,
                 z = "A curved textbox on an arbitrary path")

ggplot(df, aes(x, y, label = z)) + 
  geom_labelpath(size = 5, fill = "#F6F6FF")
```

### **`geom_textline`** and **`geom_labelline`**


- You can use `geom_textline` and `geom_labelline` as a drop in for geom_line if you want it directly labelled. - Just pass the label you want as an argument to `geom_textline` (or if you have grouped data you can pass the label as an aesthetic mapping).


**Add text to the end of the line**

```{r}
ggplot(pressure, aes(temperature, pressure)) +
  geom_textline(label = "Mercury vapor pressure", size = 6,
                linewidth = 1, linecolor = "red4", linetype = 2, 
                color = "deepskyblue4",
                hjust = 1
                )
```


### Reference lines


- Often, a reference line added to a plot requires a text annotation. 
- We can do this directly with `geom_textabline`, `geom_textvline` and `geom_texthline`, or their text-box equivalents `geom_labelabline`, `geom_labelvline` and `geom_labelhline.` 
- Although such lines aren’t curved, there are some benefits to using the **`geomtextpath`** functions if a labelled reference line is required: only a single call is needed, co-ordinates are not required for the text label, the text can be put in-line with an appropriate break in the line automatically, and the label will orientate and curve appropriately in polar co-ordinates.


```{r}
ggplot(mtcars, aes(mpg, disp)) + 
  geom_point() + 
  geom_texthline(yintercept = 200, label = "displacement threshold", 
                 hjust = 1, color = "red4") +
  geom_textvline(xintercept = 20, label = "consumption threshold", hjust = 0.8,
                 linetype = 2, vjust = 1.3, color = "blue4") +
  geom_textabline(slope = 15, intercept = -100, label = "partition line", 
                  color = "green4", hjust = 0.6, vjust=0.4)
```



