---
title: Working with Themes and Text
date: 10-20-2023
---


```{r setup, include=FALSE, message=FALSE}
# load libraries
library(tidyverse)
library(systemfonts)
library(ggthemes)
library(ggtext)
theme_set(theme_bw())

# load data
chic <- readr::read_csv("./Data/chicago-nmmaps-custom.csv")
chic

```


# Working with Themes


Note: The `theme_*()` functions also come with several other `base_*` arguments

## Change the Overall Plotting Style

You can use built-in `ggplot2` themes, `ggthemes`, among other themes. 

In the code above I have set the default theme for my plot as theme_bw()

An example using `theme_economist`

```{r}
ggplot(chic, aes(x = date, y = temp, color = season)) +
  geom_point() +
  labs(x = "Year", y = "Temperature (°F)") +
  ggtitle("Ups and Downs of Chicago's Daily Temperatures") +
  # uses theme_economist
  theme_economist() +
  # the color is modified to scale_color_economist
  scale_color_economist(name = NULL)
```


## Change the Font of All Text Elements

All themes come with an argument called `base_family`:

More generally they have an argument like `base_*`

Note: When you set `base_family`, all the text elements will use this as their default

Here we set the font text for all the text elements to `Playfair`

```{r}
g <- ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)",
       title = "Temperatures in Chicago")

g + theme_bw(base_family = "Playfair")

```




## Change the Size of All Text Elements


```{r}
g + theme_bw(base_size = 30, base_family = "Roboto Condensed")

```


## Change the Size of All Line and Rect Elements


```{r}
g + theme_bw(base_line_size = 1, base_rect_size = 1)

```


## Creating Your Own Theme


The default is called `theme_gray`



    If you wanted to create your own custom theme, you could extract the code directly from the gray theme and modify. Note that the rel() function change the sizes relative to the base_size.


You could modify the elements for the output of `theme_gray``

```{r}
theme_gray
```

### Modifying the default theme function




```{r}
theme_custom <- function (base_size = 12, base_family = "Roboto Condensed") {
  half_line <- base_size/2
  theme(
    line = element_line(color = "black", linewidth = .5,
                        linetype = 1, lineend = "butt"),
    rect = element_rect(fill = "white", color = "black",
                        linewidth = .5, linetype = 1),
    text = element_text(family = base_family, face = "plain",
                        color = "black", size = base_size,
                        lineheight = .9, hjust = .5, vjust = .5,
                        angle = 0, margin = margin(), debug = FALSE),
    axis.line = element_blank(),
    axis.line.x = NULL,
    axis.line.y = NULL,
    axis.text = element_text(size = base_size * 1.1, color = "gray30"),
    axis.text.x = element_text(margin = margin(t = .8 * half_line/2),
                               vjust = 1),
    axis.text.x.top = element_text(margin = margin(b = .8 * half_line/2),
                                   vjust = 0),
    axis.text.y = element_text(margin = margin(r = .8 * half_line/2),
                               hjust = 1),
    axis.text.y.right = element_text(margin = margin(l = .8 * half_line/2),
                                     hjust = 0),
    axis.ticks = element_line(color = "gray30", linewidth = .7),
    axis.ticks.length = unit(half_line / 1.5, "pt"),
    axis.ticks.length.x = NULL,
    axis.ticks.length.x.top = NULL,
    axis.ticks.length.x.bottom = NULL,
    axis.ticks.length.y = NULL,
    axis.ticks.length.y.left = NULL,
    axis.ticks.length.y.right = NULL,
    axis.title.x = element_text(margin = margin(t = half_line),
                                vjust = 1, size = base_size * 1.3,
                                face = "bold"),
    axis.title.x.top = element_text(margin = margin(b = half_line),
                                    vjust = 0),
    axis.title.y = element_text(angle = 90, vjust = 1,
                                margin = margin(r = half_line),
                                size = base_size * 1.3, face = "bold"),
    axis.title.y.right = element_text(angle = -90, vjust = 0,
                                      margin = margin(l = half_line)),
    legend.background = element_rect(color = NA),
    legend.spacing = unit(.4, "cm"),
    legend.spacing.x = NULL,
    legend.spacing.y = NULL,
    legend.margin = margin(.2, .2, .2, .2, "cm"),
    legend.key = element_rect(fill = "gray95", color = "white"),
    legend.key.size = unit(1.2, "lines"),
    legend.key.height = NULL,
    legend.key.width = NULL,
    legend.text = element_text(size = rel(.8)),
    legend.text.align = NULL,
    legend.title = element_text(hjust = 0),
    legend.title.align = NULL,
    legend.position = "right",
    legend.direction = NULL,
    legend.justification = "center",
    legend.box = NULL,
    legend.box.margin = margin(0, 0, 0, 0, "cm"),
    legend.box.background = element_blank(),
    legend.box.spacing = unit(.4, "cm"),
    panel.background = element_rect(fill = "white", color = NA),
    panel.border = element_rect(color = "gray30",
                                fill = NA, linewidth = .7),
    panel.grid.major = element_line(color = "gray90", linewidth = 1),
    panel.grid.minor = element_line(color = "gray90", linewidth = .5,
                                    linetype = "dashed"),
    panel.spacing = unit(base_size, "pt"),
    panel.spacing.x = NULL,
    panel.spacing.y = NULL,
    panel.ontop = FALSE,
    strip.background = element_rect(fill = "white", color = "gray30"),
    strip.text = element_text(color = "black", size = base_size),
    strip.text.x = element_text(margin = margin(t = half_line,
                                                b = half_line)),
    strip.text.y = element_text(angle = -90,
                                margin = margin(l = half_line,
                                                r = half_line)),
    strip.text.y.left = element_text(angle = 90),
    strip.placement = "inside",
    strip.placement.x = NULL,
    strip.placement.y = NULL,
    strip.switch.pad.grid = unit(0.1, "cm"),
    strip.switch.pad.wrap = unit(0.1, "cm"),
    plot.background = element_rect(color = NA),
    plot.title = element_text(size = base_size * 1.8, hjust = .5,
                              vjust = 1, face = "bold",
                              margin = margin(b = half_line * 1.2)),
    plot.title.position = "panel",
    plot.subtitle = element_text(size = base_size * 1.3,
                                 hjust = .5, vjust = 1,
                                 margin = margin(b = half_line * .9)),
    plot.caption = element_text(size = rel(0.9), hjust = 1, vjust = 1,
                                margin = margin(t = half_line * .9)),
    plot.caption.position = "panel",
    plot.tag = element_text(size = rel(1.2), hjust = .5, vjust = .5),
    plot.tag.position = "topleft",
    plot.margin = margin(rep(base_size, 4)),
    complete = TRUE
  )
}
```


- You can only overwrite the defaults for all elements you want to change. 
- However, here we listed all to illustrate that you can change literally change everything!



#### Applying our theme

```{r}
theme_set(theme_custom())

ggplot(chic, aes(x = date, y = temp, color = season)) +
  geom_point() + labs(x = "Year", y = "Temperature (°F)") + guides(color = "none")
```

**This way of changing the plot design is highly recommended!** 

It allows you to quickly change any element of your plots by changing it once. 

You can within a few seconds plot all your results in a congruent style and adapt it to other needs (e.g. a presentation with bigger font size or journal requirements).

## Update the Current Theme

You can also set quick changes using theme_update():


```{r}
theme_custom <- theme_update(panel.background = element_rect(fill = "gray60"))

# we have updated the panel background

ggplot(chic, aes(x = date, y = temp, color = season)) +
  geom_point() + labs(x = "Year", y = "Temperature (°F)") + guides(color = "none")
```

Update other elements too


```{r}
theme_custom <- theme_update(
  panel.background = element_rect(fill = "white"),
  panel.grid.major = element_line(linewidth = .5),
  panel.grid.minor = element_blank()
)
```




# Working With Text

## Add Labels to Your Data

```{r}
set.seed(2020)

sample <- chic |>
  dplyr::group_by(season) |>
  dplyr::sample_frac(0.01)

## code without pipes:
## sample <- sample_frac(group_by(chic, season), .01)

ggplot(sample, aes(x = date, y = temp, color = season)) +
  geom_point() +
  geom_label(aes(label = season), hjust = .5, vjust = -.5) +
  labs(x = "Year", y = "Temperature (°F)") +
  xlim(as.Date(c('1997-01-01', '2000-12-31'))) +
  ylim(c(0, 90)) +
  theme(legend.position = "none")
```


You will notice that there is overlap in the above chart

To avoid overlap you can use `ggrepel` as follows:


```{r}
library(ggrepel)

ggplot(sample, aes(x = date, y = temp, color = season)) +
  geom_point() +
  geom_label_repel(aes(label = season), fontface = "bold") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(legend.position = "none")
```


It may look nicer with filled boxes so we map `season` to `fill` instead to `color` and set a white color for the text:



```{r}
ggplot(sample, aes(x = date, y = temp)) +
  geom_point(data = chic, size = .5) +
  geom_point(aes(color = season), size = 1.5) +
  geom_label_repel(aes(label = season, fill = season),
                   color = "white", fontface = "bold",
                   segment.color = "grey30") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(legend.position = "none")
```



## Add Text Annotations

- There are several ways how one can add annotations to a ggplot. 
- We can again use annotate(geom = "text"), annotate(geom = "label"), geom_text() or geom_label():


```{r}
g <-
  ggplot(chic, aes(x = temp, y = dewpoint)) +
  geom_point(alpha = .5) +
  labs(x = "Temperature (°F)", y = "Dewpoint")

g + 
  annotate(geom = "text", x = 25, y = 60, fontface = "bold", 
                label = "Dewpoint - Temperature Annotation")
```

#### Using `geom_text`

However, now ggplot has drawn one text label per data point—that’s 1,461 labels and you only see one! 

You can solve that by setting the `stat` argument to `"unique"`:

```{r}
g +
  geom_text(aes(x = 25, y = 60,
                label = "Dewpoint - Temperature Annotation"),
            ## see only the text-annotation
            stat = "unique")
```



## Use Markdown and HTML Rendering for Annotations


```{r}

lab_md <- "This plot shows **temperature** in *°F* versus **ozone level** in *ppm*"

g +
  geom_richtext(aes(x = 35, y = 3, label = lab_md),
                stat = "unique")
```




```{r}
lab_html <- "&#9733; This plot shows <b style='color:red;'>temperature</b> in <i>°F</i> versus <b style='color:blue;'>ozone level</b>in <i>ppm</i> &#9733;"

g +
  geom_richtext(aes(x = 33, y = 3, label = lab_html),
                stat = "unique")
```



    The other geom from the {ggtext} package is geom_textbox(). This geom allows for dynamic wrapping of strings which is very useful for longer annotations such as info boxes and subtitles.
    

```{r}
lab_long <- "**Chicago Temperature Data**<br><i style='font-size:8pt;color:red;'>Chicago temperature and dewpoint. The data shows Chicago Temperature and dewpoint. Dewpoint and Temperature relationship in Chicago<br>.Dewpoint and Temperature relationship in Chicago</i>"

g +
  geom_textbox(aes(x = 40, y = 10, label = lab_long),
               width = unit(15, "lines"), stat = "unique")

```

Note that it is not possible to either rotate the textbox (always horizontal) nor to change the justification of the text (always left-aligned).








